---
title: "Lab 3: Exploring Happiness - Team Report"
author: "Kelsey McInturff"
date: "10/3/2019"
output: html_document
---

## Background

The General Social Survey (GSS) has been run by NORC every other year since 1972 to keep track of current opinions across the United States. We are working with an excerpt of the dataset that covers years 1972 to 2016. The variable names along with a brief description are as follows: 

| Variable      | Description   |    
| ------------- |:-------------:| 
| Year          | Year for this respondent | 
| Age           | Age of respondent | 
| Sex           | Sex of respondent     | 
|....           | .....      | 

Now it's important to understand what each column stands for. You can find a codebook with explanations for each of the variables at https://gssdataexplorer.norc.org/

## Part 1: Data Cleaning 

```{r include = FALSE}
hap <- readRDS(choose.files())

names(hap) = tolower(names(hap))
```


Initial exploration revealed that some data processing was needed before we could proceed with further visualization. All missing values were encoded with 'NA' as follows:
```{r}
library(dplyr)
happydf<-data.frame(hap)
happydf<-na_if(happydf,"IAP")
happydf<-na_if(happydf,"DK")
happydf<-na_if(happydf,"NA")
head(happydf)

sum(is.na(happydf))
```

In total there were 36391 number of missing values for the following 11 variables. 

Variables were also cast into the correct type in order to ease data visualization. This was done by:
```{r}
sapply(happydf,class)
cols<-c("happy","sex","marital","degree","finrela","health","partyid","polviews")
happydf[cols]<-lapply(happydf[cols], factor)
sapply(happydf,class)
happydf$age[happydf$age=="89 OR OLDER"]<-89
happydf$age=as.numeric(happydf$age)
sapply(happydf,class)
```

Factors were also organized into meaningful order as follows: 
```{r}
library(plyr)
happydf<-arrange(happydf,year,age,sex,marital,degree,finrela,health,partyid,polviews)
head(happydf)
```


## Part 2: Data Exploration 

Two variables of interest are Happiness Levels and Degree. Their distributions are as follow: 
```{r}
library(ggplot2)
ggplot(data=happydf,aes(happydf$happy))+geom_bar(fill="yellow",stat="count")+labs(x="Happiness Level",y="Count",title="Bar Plot of Happiness Levels")+theme(axis.text.x = element_text(angle = 90))

```
The distibtion for the bar plot of happiness is close to symmetrical, suggesting that many people describe themselves as pretty happy, whereas less people indicate that they are at the extremes of not too happy or very happy.


```{r}
ggplot(data=happydf,aes(x=factor(degree,level=c('NA','LT HIGH SCHOOL','HIGH SCHOOL','JUNIOR COLLEGE','BACHELOR','GRADUATE'))))+geom_bar(fill="red",stat="count")+labs(x="Degree Type",y="Count",title="Bar plot of Education Degrees Recieved")+theme(axis.text.x = element_text(angle = 90))
```
The distribution for the bar plot of education degrees recieved is right skewed, indicating that is is much more common for a person to have lower degrees like a high school diploma than a higher degree like a graduate degree.


```{r}
ggplot(happydf,aes(x=factor(degree,level=c('NA','LT HIGH SCHOOL','HIGH SCHOOL','JUNIOR COLLEGE','BACHELOR','GRADUATE')),fill=happy))+geom_bar(position="fill")+theme(axis.text.x = element_text(angle = 90))+labs(x="Degree Type",y="Count",title="Happiness by Degree Type")

ggplot(happydf,aes(health,fill=happy))+geom_bar(position="fill")+theme(axis.text.x = element_text(angle = 90))+labs(x="Health Condition",y="Count",title="Happiness by Health")

ggplot(happydf,aes(x=factor(marital,level=c('NEVER MARRIED','MARRIED','SEPARATED','DIVORCED','WIDOWED')),fill=happy))+geom_bar(position="fill")+theme(axis.text.x = element_text(angle = 90))+labs(x="Marital Status",y="Count",title="Happiness by Marital Status")
```

Using the happiness data, the relationship between level of happiness and a few other variables was explored. From this exploration, some correlations can be noted. From the plot of happiness and degree type above, there is a gradual increase of general happiness (pretty happy and very happy frequency) the level of the degree earned gets higher. For the plot comparing happiness and health condition, there is an increase in the number of not too happy individuals as health condition declines to poor. For the plot of happiness and marital status, there is less of a direct correlation. Rather, it seems only that married people reported being very happy more than others. The other groups in the marital staus variable have similar distributions of happiness. Overall, it seems that happiness is positively correlated with degree level, positively correlated with health, and related somewhat with being married.


